---
title: "Final Project Spring '19"
author: "Asya Kirgiz"
date: "April 28, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Birth Data Description

This data set contains monthly live births in thousands for the United states from year 1948 to 1979:
```{r birth}
data(birth,package="astsa")
start(birth)
end(birth)
frequency(birth)

summary(birth)
```

Data references: <http://www.stat.pitt.edu/stoffer/tsa4/>.

## Plots

Plots help you visualize the data and spot any trends:

```{r plot}
plot(birth)
abline(reg=lm(birth~time(birth)))
cycle(birth)
boxplot(birth ~ cycle(birth))
```

## Investigate seasonal patterns in the data

Notice seasonal difference between number of births in Aprils vs. Octobers

```{r invest, echo=FALSE}
plot(aggregate(birth)) #removes seasonal changes
birth.April <- window(birth, start=c(1948,4), freq = TRUE)
plot(birth.April)
birth.October <- window(birth, start=c(1948,10), freq = TRUE)
plot(birth.October)
```

## Decompose using multiplicative model


```{r decomp}
birth.decom <- decompose(birth, type="mult")
plot(birth.decom)
```

The trend over years seem to folow a smoth curve that is showing an upward trend until 1962, then turning and displaying a downward trend until 1975 and then reversing into an upward trend again. These changes in number of births in US could be explained by historical and econamical events.
There are clear seasonal variations in the data. 
The random part of the data does look random.

## Transformation

```{r trans}
qqnorm(birth)
```
Examining Q-Q norm plot of birth data which looks close to a straight line suggests that the distribution could be reasonably assumed to be normal.

## Autocorrelation for lag 1, lag 2, lag 3 

```{r corr}
acf(birth)$acf[2]
acf(birth)$acf[3]
acf(birth)$acf[4]
pacf(birth)
```

## Model fitting

This naive model is actually a decent fit judging from graphs below:
```{r naive, echo=FALSE}
library(forecast)
birth_fit = naive(birth)
plot(birth_fit)
```

Let's look at autocorrelation of birth differences from month to month:
```{r diff, echo=FALSE}
birth_change <- diff(birth) #changes in birth rate
acf(birth_change, lag.max = 3)
```

Now, let's fit a first order moving average model MA_birth_change, a first order auto-regression model AR_birth_change and an ARIMA model.
```{r models}
MA_birth_change <- arima(birth_change, order=c(0,0,1))
AR_birth_change <- arima(birth_change, order=c(1,0,0))
ARIMA_birth_change <- arima(birth_change, order=c(1,1,1))
```


## Evaluate the models
Use residuals to diagnose the models.
```{r resid}
checkresiduals(birth_fit)
checkresiduals(MA_birth_change)
checkresiduals(AR_birth_change)
checkresiduals(ARIMA_birth_change)

accuracy(birth_fit)
accuracy(MA_birth_change)
accuracy(AR_birth_change)
accuracy(ARIMA_birth_change)
```
The residuals of all models resamble random distributions and their means are close to zero, therefore all of the above models are reasonable representation of birth data. ARIMA_birth_change model's residuals are closest to a normal distribution with constant variance, therefore it is the best model.

From comparing the errors of the models we can see that ARIMA_birth_change has smallest values which confirms that it is the best model.


Secondly, use information criterion AIC and BIC to evaluate the models.
```{r eval}
AIC(MA_birth_change)
AIC(AR_birth_change)
AIC(ARIMA_birth_change)

BIC(MA_birth_change)
BIC(AR_birth_change)
BIC(ARIMA_birth_change)
```
The ARIMA_birth_change model has lowest AIC and BIC than others, so we'll use it for foreasting.


## Forecasting

Predict birth numbers one year into the future:
```{r forec}
predict(ARIMA_birth_change, n.ahead = 12)
future = forecast(ARIMA_birth_change, h = 12)
plot(future)
```
